---
title: "Assignment 1 - Introduction and Wrangling"
author: 'Nafis Mohammad'
date: 'October 06, 2026'
output:
  pdf_document: default
  html_document: default
---

## Set-Up

The dataset can be accessed here: https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29

The phenotype attributions are as follows: 

1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

Here is some code to read in the dataset

> bCancer <- read.csv(here("3_csv", "breastCancer.csv"), header=T)

## Questions

1. When installing a package with the `install.packages()` function, the incoming package name needs to be enclosed in quotations. Why is this? [1]

```{r}
install.packages("here", repos = "https://cloud.r-project.org")
install.packages("tidyverse", repos = "https://cloud.r-project.org")
library(here)
library(tidyverse)
```

The incoming package name needs to be enclosed in quotations because its input is valued as a character string, not as an active variable or symbol within the work space.

2. Create a vector object with a numeric value, a character value, and a logical value. What data type is this vector and why? [1]

```{r}
x <- c(44, "Lion", TRUE)
x

```
The data type as seen above is a character vector, the reason it does this is because R performs type coercion and converts every value to the most flexible type which in this statement here is character.

3. Create a new object called `recent_fruits`.

a) Create 5 rows with the rownames as different fruits and 3 columns named "colour", "shape", and "taste". Describe the fruit's colour and shape in one word, rank the taste out of 10. Fill out the table! [0.5]

```{r}
recent_fruits <- data.frame(
  colour = c("red","yellow","orange","green","black"),
  shape = c("round","curved","round","oval","pebbled"),
  taste = c(10,10,8,6,4),
  row.names = c("Apple","Banana","Orange","Kiwi","Blackberry")
)

recent_fruits
```


b) Calculate the mean taste rank from your previous table. [0.5]

```{r}
mean_taste <- mean(recent_fruits$taste)
mean_taste
```


c) Write a function called "middle_mean" that removes the largest and smallest taste rating, and then calculates the mean. Use the function on your dataset. *Hint*: you may need any of these functions: min(), max(), nrow(), length(). [0.75]

```{r}
middle_mean <- function(x) {
  x <- x[x != min(x) & x != max(x)]
  mean(x)
}

middle_mean(recent_fruits$taste)
```


4. Use the Cancer data set.

```{r}
breastCancer <- read.csv("breastCancer.csv")
```


a) Subset the samples with a mean radius less than 20 into a new object called `large_tumour`. [0.5]

```{r}
large_tumour <- breastCancer[breastCancer$radius_mean <20,]
large_tumour
```


b) From the `large_tumour` object, subset for the columns describing the `id`, `diagnosis`, `radius_mean`, `texture_mean`, `smoothness_mean`, and `compactness_mean`. Reshape the data into a longer format so that there are only four columns: `id`, `diagnosis`, `variable`, and `value`. Save this into a new object called `large_tumour_long`. [0.5]

```{r}
large_tumour_long <- large_tumour %>%
  select (id,diagnosis,radius_mean,texture_mean,smoothness_mean,compactness_mean) %>%
  pivot_longer(
    cols = c(radius_mean,texture_mean,smoothness_mean,compactness_mean),
    names_to = "variable",
    values_to = "value"
  )

large_tumour_long
```


c) Group your data by diagnosis and variable, and calculate the mean, median, max, min, standard deviation, and count the number of observations. Save this into a table object called `large_tumour_summary`. *Hint*: first, figure out what type of data format you may need (`large_tumour_long` or `large_tumour`), and then check that your data columns are the correct class. [0.5]

```{r}
str(large_tumour_long)

large_tumour_summary <- large_tumour_long %>%
  group_by(diagnosis, variable) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    max = max(value),
    min = min(value),
    sd = sd(value),
    count = n()
  )

large_tumour_summary
```


d) Which patients have the 5 largest `radius_mean`, and what is their diagnosis? Which patients have the 5 smallest `radius_mean`, and what is their diagnosis? What does this trend potentially tell you about the data? *Hint*: there are multiple measurements for each patient, so use mean() to average these measurements. [0.75]

```{r}
patient_radius <- breastCancer %>%
  group_by(id,diagnosis) %>%
  summarize(
    mean_radius = mean(radius_mean)
  )


largest_radius <- patient_radius %>%
  arrange(desc(mean_radius)) %>%
  head(5)

largest_radius


smallest_radius <- patient_radius %>%
  arrange(mean_radius) %>%
  head(5)

smallest_radius
```

